Tree growth, mortality, and climatic stress in west-coast states, USA

### 2025-11-19 Manuscript.Rmd
Latest manuscript changes to RMD file

grow.tot.sig <- grow.AvB.sig + grow.A.sig + grow.B.sig

@@

mort.tot.sig <- mort.AvB.sig + mort.A.sig + mort.B.sig

@@

fig.domains <- fig_nums(name = "fig_domains",
                        caption = "Decadal (A) growth and (B) mortality domain estimates for *Calocedrus decurrens*.  The six domains are defined as FIA plots that fall [A]bove or [B]elow the CWD change threshold with [L]ow, [M]edium, or [H]igh initial CWD values (see text for details). E.g., 'AL' is Above CWD change threshold, Low initial CWD value. The distribution of plot CWD change and initial values (C) is presented as well, with the blue horizontal line indicating the CWD change threshold and the orange vertical lines representing the quantile boundaries between Low, Medium, and High initial CWD values." )

@@

@@

Numerous tree species in western North America exhibit increased mortality rates or decline.  van Mantgem et al. (2009) modeled mortality rates for the genera *Abies* and *Pinus* in the western USA as increasing from less than 0.05% annually in 1960 to almost 0.2% by 2010. Johnston et al. (2025) documented a 29% 10-year mortality rate of 300+ yr old trees in Eastern Oregon.  Perret et al. (2023) estimated a decadal range-wide mortality rate of 25% for Subalpine fir (*Picea engelmannii*).  High-severity fires in California in 2020 and 2021 killed 13%-19% of extant giant sequoia (*Sequoiadendron gianteum*) with >1.07-m bole diameters (Stephenson and Brigham 2021, Shive et al. 2022); recruitment in these burns was depressed (Soderberg et al. 2022).   Five of eight common tree species in western North America exhibit declines as measured by relative stand density (Stanke et al. 2021).  Shriver et al. (2022) found that four of five juniper (*Juniperus* spp.) and piñon pine (*Pinus* spp.) species were declining in portions of their range, comparing mortality to recruitment rates.

Climate change as manifest through increases or intensification in drought and aridity can affect tree mortality directly through physiological damage as well as indirectly by increasing wildfire frequency and severity and by promoting insect and disease outbreaks.  Hammond et al. (2022) found globally that the intensity of drought and heat events are directly associated with widespread forest die-off events.  Wildfire frequency and area burned in the western United States has increased in concert with the longer fire seasons, hotter summer temperatures, and earlier snowmelt (Abatzoglou and Williams 2016, Westerlin 2016, Juang et al. 2022).  There has also been a regional increase in the area of bark beetle (subfamily *Scolytinae* spp.) outbreaks (Meddens et al. 2012, Hicke et al. 2016).  Many of these indirect climate-influenced mortality causes may be attributable to the physical condition of trees following abnormal or prolonged exposure to water stress.

Carbon starvation and hydraulic failure can drive drought-induced mortality (see reviews by Adams et al. 2010 and McDowell et al. 2008).  Carbon starvation occurs when isohydric species (e.g., piñon pine, *Pinus edulus*; Breshears et al. 2009), or those which mainly rely on stomatal closure to avoid excessive water loss through evapotranspiration (ET), receive insufficient atmospheric carbon dioxide for photosynthesis.  Anisohydric plants, or those who do not strongly regulate water loss through stomatal closure, risk hydraulic failure if sufficient xylem cavitation occurs due to insufficient water uptake needed to meet ET demands (e.g., as seen in *Populus tremuloides*; Anderegg et al. 2015). Adams et al. (2017) performed a review of drought-induced mortality across 26 species and concluded that hydraulic failure occurred broadly across taxa while non-structural carbohydrate reserves at mortality differed across taxa.  Regardless of mechanism, drought (specifically, water deficit) appears to be an important direct driver of mortality for tree taxa (Hammond et al. 2022).

@@

One regionally useful metric of water stress is Climatic Water Deficit (CWD; Stephenson 1998).  Stephenson (1990) defines CWD as evaporative demand not met by available water, or the difference between plant water demand (actual evapotranspiration, AET), and plant water supply (potential evapotranspiration); these variables are driven by temperature and precipitation, respectively.  The Anderegg  et al. (2015) tree mortality study relied on CWD to capture drought occurrence.  Das et al. (2013) determined that low-elevation California tree mortality was best modeled using CWD.  Restaino et al. (2016) found decreased Douglass fir growth with an increase in CWD.

@@

Climatic data such as mean annual temperature and CWD can be combined with FIA data to obtain climate-related design-based estimates. Monleon and Linz (2015) evaluated the evidence of tree species range shifts by subtracting the design-based mean estimates of FIA plot temperatures for seedlings from those of trees. FIA plot revisit data were mostly incomplete at the time of their analysis.  Groom and Monleon (2023) used a mostly-completed revisit data set to compare plot climate variable values (vapor pressure deficit [VPD], temperature, and precipitation) to estimate shifts in plot occupancy and density.  Both studies used a single climate variable value associated with each plot for estimation; e.g., Groom and Monleon (2023) examined shifts in species density by estimating the difference in VPD means between plots that gained trees of a given species and plots that lost trees.  They documented, however, that the initial climate variable values could (and often did) change between plot visits, with species' plots often exhibiting positive or negative correlations between the initial visit values and between-visits change value of the climate variable.  They identified that their design-based analysis would potentially improve if they could simultaneously include both initial visit and change values for the climate variable.

The purpose of this analysis is to demonstrate an approach for a bivariate design-based estimation procedure. We estimate tree species’ growth and mortality across the western U.S. states of California, Oregon, and Washington as they relate to plot-level initial-visit water stress (initial CWD) and the change in water stress between plot visits (delta CWD).  We approached this bivariate analysis by dividing species’ plot data into domains based on initial and delta CWD values, finding growth and mortality estimates for each domain, and estimating domain contrasts to establish differences between domains.  Secondarily we provide decadal growth and mortality estimates for our analysis species overall and by state to establish species status and to inform our CWD analysis findings.

@@

Plots in the three states are sampled every 10 years with a target of annually sampling a panel of 10% of plots every year (Thompson 2015). Our analysis depends on examining the fate and growth of trees measured during two sequential visits (referred to as the first and second plot visit), ideally from a complete sample panel remeasure.  We obtained our FIA data from the USDA FIA DataMart web portal (https://research.fs.usda.gov/products/dataandtools/fia-datamart; downloaded May 5, 2025).  The database included plot data sampled from 2001 to 2023.

@@

We obtained plot-level CWD data from the model TerraClimate (Abatzoglou et al. 2018) for June, July, and August (summer months) between 1980 and 2023. Latitude and longitude were based on FIA-fuzzed coordinates.  We defined the 'initial CWD' values as the 20-year mean of summer CWD values for each plot prior to and including the first visit year.  We then found the mean of summer CWD values between visits and including the second visit year.  We then defined 'delta CWD' values as the difference between this second value and initial CWD value.  Figure 1 presents maps of both the initial CWD and delta CWD values for forested and nonforested FIA plots.

@@

We estimated, for each species examined, six domains based on CWD values (Figure 2C).   The initial CWD plot values are divided into three species-specific quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High).  Delta CWD plot values for all species were divided into those above a 5mm CWD change threshold (Above) and those below the threshold (Below). The 5mm threshold was selected because species' delta CWD value distributions did not differ greatly and because 5mm is approximately the mean delta CWD value across species (Table S2).  The Below domains may includes negative delta CWD values (i.e., mean CWD was larger prior to the initial visit than between visits).  The domain designations AL, AM, and AH correspond to Above Low, Above Medium, and Above High. The Below domains are similarly labeled (BL, BM, BH). We also report species mortality and growth estimates for all plots across and within Oregon, Washington, and California irrespective of CWD values.  We limited the number of species examined to those with more than 400 occupied plots across the three states.  We calculated estimates for all CWD-based domains with 10 or more plots.

@@

Our analysis approach results in many comparisons.  We conducted a Monte Carlo (MC) permutation analysis to provide species-level false positive rates (p-values) for "significant" contrasts. For each of 1000 MC permutation analysis iterations the code estimated each all species' bootstrap confidence intervals as described above. The MC permutation analysis sampled each species' plot domain assignments without replacement.  Given this sampling approach, and since species distributions were not independent, the p-values are strictly interpretable at the species level (e.g., we cannot speak to the likelihood of observing a specific number of significant findings for a contrast type). The distance of MC permutation contrast CI bounds from zero were compared against test CI bounds. The MC permutation procedure recorded a null significant value if the absolute values of the permutation iteration CI bounds were at least as far from zero as observed bounds.

@@

Our study tracked the fate of `r comma(analysis.stats$tree.num)` trees from `r length(sel.spp)` species in `r comma(analysis.stats$plot.num)` plots that were visited twice by FIA crews. Of the `r length(sel.spp)` species, `r analysis.stats$n.gym.spp` were gymnosperms.  The two visits were 10 years apart for `r pct.remeas[2]`% of plots (`r pct.remeas[1]`% were 9 years, `r pct.remeas[3]`% were 11 years). Plots were distributed in `r analysis.stats$strat.num` strata.

Mean overall growth rates for species varied between `r min.grow.spp$mean` (*`r min.grow.spp$spp`*) and `r max.grow.spp$mean` (*`r max.grow.spp$spp`*, `r t.ref("tbl_growth")`) cm$^2$/decade.  Six species had greater growth (non-overlapping CIs) in California than in Oregon and/or Washington: *Abies concolor*, *Abies grandis*, *Pinus ponderosa*, *Tsuga heterophylla*, *Arbutus menziesii*, and *Notolithocarpus densiflorus*.  Three species had greater growth in Oregon (*Calocedrus decurrens*, *Picea engelmanii*, and *Pseudotsuga menziesii*) and two species had greater growth in Washington (*Abies amablis* and *Pinus monticola*).  Decadal mean mortality rates varied between `r min.mort.spp$mean` (*`r min.mort.spp$spp`*) and `r max.mort.spp$mean` (*`r max.mort.spp$spp`*, `r t.ref("tbl_mort")`). Four species in California exhibited greater mortality rates than in Oregon and/or Washington: *Abies concolor*, *Pinus ponderosa*, *Acer macrophyllum*, and *Quercus kelloggii*.  One species had a greater mortality rate in Oregon (*Tsuga mertensiana*) and Washington had five (*Abies grandis*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, and *Pseudotsuga menziesii*).  Seven species had overall mortality rates $\geq$ 0.20 (*Abies concolor*, *Abies lasiocarpa*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Alnus rubra*, and *Quercus chryselopis*) while at a state level seven species had mortality rates $\geq$ 0.25 (California: *Acer macrophyllum* and *Alnus rubra*; Oregon: *Abies lasiocarpa*; Washington: *Abies grandis*, *Picea engelmannii*, *Pinus contorta*, and *Pinus monticola*).

@@

Individual species growth and mortality domain patterns could differ (e.g., Figures `r fig_nums("fig_domains", display = "num")`A and `r fig_nums("fig_domains", display = "num")`B). Growth domain contrasts (Figure `r fig_nums("fig_diffgrow", display = "num")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to result in substantially more "significant" negative or positive estimates, indicating that species overall did not exhibit more or less growth in the Above, Low, or High domains. All three columns of Figure `r fig_nums("fig_diffgrow", display = "num")` had over 20 "significant" contrasts. Depending on the species, growth may have been susceptible (negatively or positively) to a change in CWD and/or differed by initial CWD values. *Abies amabilis*, *Abies grandis*, *Calocedrus decurrens*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Alnus rubra* exhibited greater growth in the Low domains than the Medium and/or High domains.  Except for *Alnus rubra*, these results predominantly occurred in both the Below domain (2 spp.) or the Above and Below domains (5 spp.).  *Thuja plicata* had greater growth in the Above and Below Medium domains than either High or Low domains. *Abies concolor*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Arbutus menziesii*, and *Quercus kelloggii* had greater growth in the more water stressed High domains than the Medium and/or Low domains. Except for *Larix occidentalis*, these results predominantly occurred in both the Above domain (3 spp.) or the Above and Below domains (3 spp.). Other species exhibited no 'significant' findings (4 spp.) or displayed other result patterns (3 spp.).

Mortality domain contrasts are similar in certain regards to the growth contrasts: results for the three types of domain contrasts did not appear to be predominantly positive or negative (Figure `r fig_nums("fig_diffmort", display = "num")`, `r t.ref("tbl_summ_grow_mort")`).  There were fewer mortality estimates with "significant" findings than growth (`r mort.tot.sig` vs `r grow.tot.sig`). There were fewer differences for "Above vs. Below Threshold" contrasts (`r mort.AvB.sig`, `r t.ref("tbl_summ_grow_mort")`) than for Above (`r mort.A.sig`) or Below (`r mort.B.sig`), indicating that mortality may not be as strongly associated with delta CWD as by the initial CWD value.  *Larix occidentalis*, *Picea engelmannii*,	*Pinus contorta*,	*Pinus monticola*, *Thuja plicata*,	and *Arbutus menziesii* had greater mortality in the lower water deficit Low domains than Medium and/or High domains. These results occurred in Above (3 spp.) or both Above and Below (3 spp.) domains. Mortality rates for *Abies lasiocarpa*,	*Calocedrus decurrens*,	*Pinus lambertiana*, *Pinus ponderosa*,	*Pseudotsuga menziesii*,	*Tsuga heterophylla*,	*Tsuga mertensiana*,	*Acer macrophyllum*,	and *Notolithocarpus densiflorus* were greater in the High domains than the Medium and/or Low domains.  These species were approximately evenly split between differences being present in the Above (4 spp.) or Below (5 spp.) domains, but none with "significant" findings in both the Above and Below domains.  Five species had no significant contrasts.

The dashboard revealed domains of interest ("significant") for certain species that aligned with geographical regions.  For the Above and Below Low domains *Abies lasiocarpa* had less mortality in northeastern Washington and *Tsuga mertensiana* exhibited greater growth in the same region.  *Pinus contorta* exhibited less growth and greater mortality in northern Washington (Above Low).  In the southern Sierra Nevada *Pinus lambertiana* had greater growth (Above Medium and High) and greater mortality (Above High), *Calocedrus decurrens* had greater mortality (Above High), and *Pinus monticola* had less mortality (Above High). *Acer macrophyllum* exhibited less growth in the Sierra Nevada and southern Klamath Mountains (Above High) and greater growth in northwestern Washington in the Coast and Cascade ranges (Above and Below Low).

@@

Across the 23 species examined we found no clear pattern of response to initial and/or delta CWD values.  Responses were species-specific, potentially driven in some cases by local phenomena and captured by a domain category or categories.  We did find more “significant” CWD contrasts for species growth than mortality, possibly indicating that growth differences by CWD domain were more readily detectable than mortality over a decade, although growth and mortality findings were not necessarily negatively related.  Delta CWD increased almost everywhere between plot revisits although the magnitude of change was generally not large relative to initial CWD values.  We additionally found for some species that decadal mortality rates by state and/or region were concerningly high.

The number of "significant" Above vs. Below threshold contrasts for growth and mortality were approximately evenly distributed between positive and negative contrast estimates, with a positive difference indicating that some species exhibited higher growth or greater mortality in an Above domain (e.g., *Pinus lambertiana* for both growth and mortality) versus a Below domain (*Abies grandis* and *Pinus contorta*, respectively, for growth and mortality).  This lack of a pattern across species may indicate that the distinction of Above vs. Below is not generalizable across species, or that "significant" findings are due to confounding factors such as geographical locations with different ecological pressures.

>>>>

grow.tot.sig <- grow.AvB.sig + grow.A.sig + grow.B.sig
grow.tot <- grow.n.AvB + grow.n.A + grow.n.B

@@

mort.tot.sig <- mort.AvB.sig + mort.A.sig + mort.B.sig
mort.tot <- mort.n.AvB + mort.n.A + mort.n.B

@@

fig.domains <- fig_nums(name = "fig_domains",
                        caption = "This figure illustrates our bivariate domain approach using *Calocedrus decurrens* as an example. Here we depict decadal (A) growth and (B) mortality domain estimates.  The six domains are defined as FIA plots that fall [A]bove or [B]elow the CWD change threshold with [L]ow, [M]edium, or [H]igh initial CWD values (see text for details). E.g., 'AL' is Above CWD change threshold, Low initial CWD value. The distribution of plot CWD change and initial values (C) is presented as well, with the blue horizontal line indicating the CWD change threshold and the orange vertical lines representing the quantile boundaries between Low, Medium, and High initial CWD values." )

@@

Western North American forests are experiencing accelerating tree mortality rates in response to intensifying drought and water stress driven by climate change.  Tree mortality and growth reflect cumulative effects of direct physiological stress (xylem cavitation, carbon starvation), and indirect climate-mediated impacts (wildfire, insects, pathogens).  However, obtaining design-based estimates for these demographic rates requires addressing a methodological challenge: revisit data are necessary to measure tree fate and growth, but plots differ in their initial water stress conditions and the magnitude of change between visits, necessitating a bivariate rather than univariate estimation approach. We developed a bivariate design-based estimation framework that stratifies plots into domains defined by both baseline initial-visit climatic water deficit (CWD) and change in CWD between visits.  We applied this approach to `r comma(analysis.stats$plot.num)` National Forest Inventory plots spanning California, Oregon, and Washington, tracking `r comma(analysis.stats$tree.num)` trees from `r length(sel.spp)` species over decade-long intervals.  We estimated decadal growth and mortality overall, by state, and across six CWD domains per species, using bootstrap confidence intervals and Monte Carlo permutation tests to assess statistical significance.  Species responses were highly variable, with no universal pattern of increased mortality or reduced growth in domains experiencing elevated water stress. Several species exhibited overall mortality rates exceeding 0.20 per decade (20% decadal mortality), with state estimates of 0.34 for *Picea engelmannii* in Washington, 0.27 for *Acer macrophyllum* in California, and 0.37 for *Pinus contorta* in Washington. Domain-specific patterns aligned with known regional disturbances (bark beetles, drought, fire) for multiple species. Our bivariate framework provides a template for monitoring forest vital rates as ecosystem-level indicators of climate change response, while our species-specific findings identify or validate populations warranting immediate conservation assessment and inform forest management strategies.

@@

Western North American forests are experiencing increases in tree mortality across multiple species in concert with changing climate conditions.  Species and tree groups have incurred extraordinary losses in specific regions.  Johnston et al. (2025) documented a 29% 10-year mortality rate of 300+ yr old trees in Eastern Oregon.  High-severity fires in California in 2020 and 2021 killed 13%-19% of extant giant sequoia (*Sequoiadendron gianteum*) with >1.07-m bole diameters (Stephenson and Brigham 2021, Shive et al. 2022); recruitment in these burns was depressed (Soderberg et al. 2022).  Species and genera have also seen their mortality rates increase overall. van Mantgem et al. (2009) modeled mortality rates for the genera *Abies* and *Pinus* in the western USA as increasing from less than 0.05% annually in 1960 to almost 0.2% by 2010. Perret et al. (2023) estimated a decadal range-wide mortality rate of 25% for Subalpine fir (*Picea engelmannii*).    Five of eight common tree species in western North America exhibit declines as measured by relative stand density (Stanke et al. 2021).  Shriver et al. (2022) found that four of five juniper (*Juniperus* spp.) and piñon pine (*Pinus* spp.) species were declining in portions of their range, comparing mortality to recruitment rates.

Climate change as manifest through increases or intensification in drought and aridity can affect tree mortality directly and indirectly.  Direct impacts include physiological damage while indirect causes include increasing wildfire, insect, and disease outbreak frequency and severity.  Hammond et al. (2022) found globally that the intensity of drought and heat events are directly associated with widespread forest die-off events.  Wildfire frequency and area burned in the western United States has increased in concert with the longer fire seasons, hotter summer temperatures, and earlier snowmelt (Abatzoglou and Williams 2016, Westerlin 2016, Juang et al. 2022).  There has also been a regional increase in the area of bark beetle (subfamily *Scolytinae* spp.) outbreaks (Meddens et al. 2012, Hicke et al. 2016).  Many of these indirect climate-influenced mortality causes may be attributable to the physical condition of trees following abnormal or prolonged exposure to water stress.

Carbon starvation and hydraulic failure can drive drought-induced mortality (see reviews by Adams et al. 2010, Sala et al. 2010, 2012, and McDowell et al. 2008).  Carbon starvation occurs when isohydric species (e.g., piñon pine, *Pinus edulus*; Breshears et al. 2009), or those which mainly rely on stomatal closure to avoid excessive water loss through evapotranspiration (ET), receive insufficient atmospheric carbon dioxide for photosynthesis.  Anisohydric plants, or those who do not strongly regulate water loss through stomatal closure, risk hydraulic failure if sufficient xylem cavitation occurs due to insufficient water uptake needed to meet ET demands (e.g., as seen in *Populus tremuloides*; Anderegg et al. 2015). Adams et al. (2017) performed a review of drought-induced mortality across 26 species and concluded that hydraulic failure occurred broadly across taxa while non-structural carbohydrate reserves at mortality differed across taxa.  Regardless of mechanism, drought (specifically, water deficit) appears to be an important direct driver of mortality for tree taxa (Hammond et al. 2022).

@@

Given the importance of water stress on tree mortality, selecting an appropriate metric is critical for rangewide assessments.  One regionally useful metric of water stress is Climatic Water Deficit (CWD; Stephenson 1998).  Stephenson (1990) defines CWD as evaporative demand not met by available water, or the difference between plant water demand (actual evapotranspiration, AET), and plant water supply (potential evapotranspiration); these variables are driven by temperature and precipitation, respectively.  The Anderegg  et al. (2015) tree mortality study relied on CWD to capture drought occurrence.  Das et al. (2013) determined that low-elevation California tree mortality was best modeled using CWD.  Restaino et al. (2016) found decreased Douglass fir growth with an increase in CWD.

@@

Climatic data such as mean annual temperature and CWD can be combined with FIA data to obtain climate-related design-based estimates. Monleon and Linz (2015) evaluated the evidence of tree species range shifts by subtracting the design-based mean estimates of FIA plot temperatures for seedlings from those of trees. FIA plot revisit data were mostly incomplete at the time of their analysis.  Groom and Monleon (2023) used a mostly-completed revisit data set to compare plot climate variable values (vapor pressure deficit [VPD], temperature, and precipitation) to estimate shifts in plot occupancy and density.  Both studies used a single climate variable value associated with each plot for estimation; e.g., Groom and Monleon (2023) examined shifts in species density by estimating the difference in VPD means between plots that gained trees of a given species and plots that lost trees.

Groom et al. (2023) documented, however, that the initial climate variable values could (and often did) change between plot visits, with species' plots often exhibiting positive or negative correlations between the initial visit values and between-visits change value of the climate variable.  They identified that their design-based analysis would potentially improve if they could simultaneously include both initial visit and change values for the climate variable.

Our study has two objectives.  The first is to demonstrate, across 23 species, a bivariate design-based estimation approach for tree growth and mortality that simultaneously accounts for initial climatic conditions and change between measurements.  This addresses a gap in design-based analyses of forest change.  Second, we provide current growth and mortality estimates for our analysis species across our study area and by state to establish species status and to inform our climate analysis findings.

@@

Plots in the three states are sampled every 10 years with a target of annually sampling a panel of 10% of plots every year (Thompson 2015). Our analysis depends on examining the fate and growth of trees measured during two sequential visits (referred to as the first and second plot visit), ideally from a complete sample panel remeasure. Tree growth is the observed increase in marked tree bole diameter between visits (cm$^2$) while mortality is captured from the change in tree condition. We obtained our FIA data from the USDA FIA DataMart web portal (https://research.fs.usda.gov/products/dataandtools/fia-datamart; downloaded May 5, 2025).  The database included plot data sampled from 2001 to 2023.

@@

We obtained plot-level CWD data from the model TerraClimate (Abatzoglou et al. 2018) for June, July, and August (summer months) between 1980 and 2023. Latitude and longitude were based on FIA-fuzzed coordinates.  We defined the 'initial CWD' values as the 20-year mean of summer CWD values for each plot prior to and including the first visit year (initial CWD = mean(summer CWD from visit year$_{-19}$ to year$_0$).  We then found the mean of summer CWD values between visits and including the second visit year (second visit CWD = mean(summer CWD from year$_1$ to year$_10$) .  We then defined delta CWD = second visit CWD - initial CWD.  Figure 1 presents maps of both the initial CWD and delta CWD values for forested and nonforested FIA plots.

@@

We estimated, for each species examined, six domains based on CWD values.  The domains allow us to parse the plot CWD values into three levels for initial CWD values and two levels for delta CWD values.  We created the domains based on initial CWD plot values by dividing the values into three species-specific quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High; Figure 2C).  We defined the two domains for delta CWD plot values for all species as those above a 5mm CWD change threshold (Above) and those below the threshold (Below). The 5mm threshold was selected because species' delta CWD value distributions did not differ greatly and because 5mm is approximately the mean delta CWD value across species (Table S2).  The Below domains may includes negative delta CWD values (i.e., mean CWD was larger prior to the initial visit than between visits).  The domain designations AL, AM, and AH correspond to Above Low, Above Medium, and Above High. The Below domains are similarly labeled (BL, BM, BH). Parsing the plot data into six domains allow for bivariate examinations of each species' data.  This bivariate approach allows us to ask, for example, whether trees in historically dry sites (High initial CWD) respond differently to increased water stress (Above threshold delta CWD) compared to trees in mesic sites (Low initial CWD) experiencing similar increases. We calculated estimates for all CWD-based domains with 10 or more plots.

We additionally report species mortality and growth estimates for all plots across and within Oregon, Washington, and California irrespective of CWD values.  We limited the number of species examined to those with more than 400 occupied plots across the three states.

@@

Our analysis approach results in hundreds of comparisons.  We conducted a Monte Carlo (MC) permutation analysis to provide species-level false positive rates (p-values) for "significant" contrasts. For each of 1000 MC permutation analysis iterations the code estimated each all species' bootstrap confidence intervals as described above. The MC permutation analysis sampled each species' plot domain assignments without replacement.  Given this sampling approach, and since species distributions were not independent, the p-values are strictly interpretable at the species level (e.g., we cannot speak to the likelihood of observing a specific number of significant findings for a contrast type). The distance of MC permutation contrast CI bounds from zero were compared against test CI bounds. The MC permutation procedure recorded a null significant value if the absolute values of the permutation iteration CI bounds were at least as far from zero as observed bounds.

Given the large number of contrasts conducted (`r mort.tot` each for growth and mortality), we adopted a conservative approach to interpretation. Aside from discussing patterns of findings across species, we focus discussion on contrasts where (1) the 95% CI does not include zero, and (2) the MC permutation test yields p < 0.001, indicating that the observed pattern is unlikely under random domain assignment.  Contrasts with less stringent p-values are shown in figures but are generally not discussed individually as they may represent Type I errors.

@@

Our study tracked the fate of `r comma(analysis.stats$tree.num)` trees from `r length(sel.spp)` species in `r comma(analysis.stats$plot.num)` plots that were visited twice by FIA crews. Of the `r length(sel.spp)` species, `r analysis.stats$n.gym.spp` were gymnosperms.  The two visits were 10 years apart for `r pct.remeas[2]`% of plots (`r pct.remeas[1]`% were 9 years, `r pct.remeas[3]`% were 11 years). Plots were distributed across `r analysis.stats$strat.num` strata spanning forested regions in all three states, from coastal temperate rainforest to semi-arid interior woodlands.

Mean overall growth rates for species varied between `r min.grow.spp$mean` (*`r min.grow.spp$spp`*) and `r max.grow.spp$mean` (*`r max.grow.spp$spp`*, `r t.ref("tbl_growth")`) cm$^2$/decade.  Six species had greater growth (non-overlapping CIs) in California than in Oregon and/or Washington: *Abies concolor*, *Abies grandis*, *Pinus ponderosa*, *Tsuga heterophylla*, *Arbutus menziesii*, and *Notolithocarpus densiflorus*.  Three species had greater growth in Oregon (*Calocedrus decurrens*, *Picea engelmanii*, and *Pseudotsuga menziesii*) and two species had greater growth in Washington (*Abies amablis* and *Pinus monticola*).

Decadal mean mortality rates varied between `r min.mort.spp$mean` (*`r min.mort.spp$spp`*) and `r max.mort.spp$mean` (*`r max.mort.spp$spp`*, `r t.ref("tbl_mort")`). Four species in California exhibited greater mortality rates than in Oregon and/or Washington: *Abies concolor*, *Pinus ponderosa*, *Acer macrophyllum*, and *Quercus kelloggii*.  One species had a greater mortality rate in Oregon (*Tsuga mertensiana*) and Washington had five (*Abies grandis*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, and *Pseudotsuga menziesii*).  Seven species had overall mortality rates $\geq$ 0.20 (*Abies concolor*, *Abies lasiocarpa*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Alnus rubra*, and *Quercus chryselopis*) while at a state level seven species had mortality rates $\geq$ 0.25 (California: *Acer macrophyllum* and *Alnus rubra*; Oregon: *Abies lasiocarpa*; Washington: *Abies grandis*, *Picea engelmannii*, *Pinus contorta*, and *Pinus monticola*).

@@

Individual species growth and mortality domain patterns could differ (e.g., Figures `r fig_nums("fig_domains", display = "num")`A and `r fig_nums("fig_domains", display = "num")`B). Growth domain contrasts (Figure `r fig_nums("fig_diffgrow", display = "num")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to result in substantially more "significant" negative or positive estimates, indicating that species overall did not exhibit more or less growth in the Above, Low, or High domains.

All three columns of Figure `r fig_nums("fig_diffgrow", display = "num")` had over 20 "significant" contrasts. Depending on the species, growth may have been susceptible (negatively or positively) to a change in CWD and/or differed by initial CWD values. *Abies amabilis*, *Abies grandis*, *Calocedrus decurrens*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Alnus rubra* exhibited greater growth in the Low domains than the Medium and/or High domains.  Except for *Alnus rubra*, these results predominantly occurred in both the Below domain (2 spp.) or the Above and Below domains (5 spp.).  *Thuja plicata* had greater growth in the Above and Below Medium domains than either High or Low domains. *Abies concolor*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Arbutus menziesii*, and *Quercus kelloggii* had greater growth in the more water stressed High domains than the Medium and/or Low domains. Except for *Larix occidentalis*, these results predominantly occurred in both the Above domain (3 spp.) or the Above and Below domains (3 spp.). Other species exhibited no 'significant' findings (4 spp.) or displayed other result patterns (3 spp.).

Mortality domain contrasts are similar in certain regards to the growth contrasts: results for the three types of domain contrasts did not appear to be predominantly positive or negative (Figure `r fig_nums("fig_diffmort", display = "num")`, `r t.ref("tbl_summ_grow_mort")`).  There were fewer mortality estimates with "significant" findings than growth (`r mort.tot.sig` vs `r grow.tot.sig`). There were fewer differences for "Above vs. Below Threshold" contrasts (`r mort.AvB.sig`, `r t.ref("tbl_summ_grow_mort")`) than for Above (`r mort.A.sig`) or Below (`r mort.B.sig`), indicating that mortality may not be as strongly associated with delta CWD as by the initial CWD value.

*Larix occidentalis*, *Picea engelmannii*,	*Pinus contorta*,	*Pinus monticola*, and *Thuja plicata* had greater mortality in the lower water deficit Low domains than Medium and/or High domains. These results occurred in Above (3 spp.) or both Above and Below (2 spp.) domains. Mortality rates for *Abies lasiocarpa*,	*Calocedrus decurrens*,	*Pinus lambertiana*, *Pinus ponderosa*,	*Pseudotsuga menziesii*,	*Tsuga heterophylla*,	*Tsuga mertensiana*,	*Acer macrophyllum*,	and *Notolithocarpus densiflorus* were greater in the High domains than the Medium and/or Low domains.  These species were approximately evenly split between differences being present in the Above (4 spp.) or Below (5 spp.) domains, but none with "significant" findings in both the Above and Below domains. *Arbutus menziesii* had greater mortality in Medium domains than the High domains. Five species had no significant contrasts.

The interactive dashboard revealed geographic patterns that were associated with specific domains while not necessarily captured by state-level estimates.  For the Above and Below Low domains *Abies lasiocarpa* had less mortality in northeastern Washington and *Tsuga mertensiana* exhibited greater growth in the same region.  *Pinus contorta* exhibited less growth and greater mortality in northern Washington (Above Low).  In the southern Sierra Nevada *Pinus lambertiana* had greater growth (Above Medium and High) and greater mortality (Above High), *Calocedrus decurrens* had greater mortality (Above High), and *Pinus monticola* had less mortality (Above High). *Acer macrophyllum* exhibited less growth in the Sierra Nevada and southern Klamath Mountains (Above High) and greater growth in northwestern Washington in the Coast and Cascade ranges (Above and Below Low).

@@

Analysis of 23 tree species revealed diverse, species-specific associations with climatic water deficit, with different species exhibiting contrasting relationships among growth, mortality, and water stress. Responses were potentially driven by differences in drought tolerance strategies, local disturbance regimes, and biotic interactions. We found more “significant” CWD contrasts for species growth than mortality, possibly indicating that growth differences by CWD domain were more readily detectable than mortality over a decade, although growth and mortality findings were not necessarily negatively related.  Delta CWD increased almost everywhere between plot revisits although the magnitude of change was generally not large relative to initial CWD values.  We additionally found concerningly high state and/or overall decadal mortality rates for some species.

Contrast results may be more informative at an individual species level and when considering state-level mortality and growth estimates.  Regional analyses detected several species with relatively high mortality rates including *Acer macrophyllum*, *Picea engelmannii*, *Abies lasiocarpa*, *Pinus contorta*, and *Alnus rubra*.  We do not believe our results are concerning for *Alnus rubra*.  Our estimates for this short-lived tree species align with reported background mortality rates (e.g., Brown and Courtin 2018).   However, the mortality rate for *Acer macrophyllum* in California appears to be both unreported elsewhere and substantially higher in California than in Oregon or Washington.

*Acer macrophyllum* growth was negatively associated with initial CWD values, and its growth was lower and mortality almost three times higher in California than in Washington and Oregon.  We cannot find primary literature that corroborates an increase in mortality for this species in California. McNellis et al. (2021) estimated a recent *Acer macrophyllum* overall mortality rate of 0.45%/yr (4.5%/decade) while our estimates for Washington and Oregon were twice this level, and California's estimate was six times as large.  However, Betzen et al. (2021) found widespread evidence in Washington of *Acer macrophyllum* decline (crown dieback, thinning, discoloration, reduced leaf size) and reduced growth and increased mortality in locations with higher temperatures and water limitations.  Although these results align with our contrast findings for growth, as displayed in our dashboard, they are not reflective (yet) of our state mortality findings for Washington.  We encourage a more intensive assessment of *Acer macrophyllum*’s status in California.
